---
title: "A Study on Attention Span: Testing the Effects of Caffeine and Naps"
author: "Justin Noche, Stephanie Chu"
date: "18 June 2023"
format: html
---

# Abstract

Caffeine consumption and naps are increasingly common among college students. This study examines whether caffeine and naps improve attention span. Using a **Two-Way Randomized Block Design** and **two-way ANOVA**, we analyzed a sample of islanders. Our results indicate that **Nap Interval significantly affects attention span**, while caffeine consumption and gender do not.

# Introduction

Caffeine is a stimulant found in coffee, tea, and other sources, known to increase alertness. Studies suggest that:

-   **Caffeine doses of 75 mg or more** impact attention (EFSA).
-   **Short naps of up to 30 minutes** can enhance alertness (CDC).
-   **“Caffeine naps” (caffeine before napping)** may improve focus (Centofanti et al.).

This study investigates the **combined effects of caffeine and naps** using **water, coffee, and tea** as caffeine sources and **0, 30, and 60-minute naps** as treatments.

# Experiment Design & Methods

## Participants

-   **198 participants** (99 males, 99 females) from "The Islands" virtual program.
-   Randomly assigned into equal-sized treatment groups.

## Study Design

-   **Blocking Variable:** Gender
-   **Factors:** Nap Duration (0, 30, 60 minutes), Caffeine Type (Water, Tea, Coffee)
-   **Response Variable:** Attention Score (difference before and after treatment)

```{r echo = FALSE}
# Factorial Design Representation
df = data.frame(
  Nap_Duration = factor(rep(c("0 min", "30 min", "60 min"), each=3)),
  Caffeine = factor(rep(c("Water", "Tea", "Coffee"), times=3))
)
knitr::kable(df)
```

## Procedure

1.  Recruit 198 participants, ensuring equal gender distribution.
2.  Randomly assign them to the 9 treatment groups.
3.  Conduct a **baseline attention test**.
4.  Administer assigned **nap durations and caffeine treatments**.
5.  Wait **15 minutes** after treatment.
6.  Conduct a **post-treatment attention test**.
7.  Calculate the **difference in attention scores**.

# Results

## Box Plots

The boxplots below show the distribution of attention score differences by Nap Interval and Caffeine Type:

```{r, echo=FALSE}
library(ggplot2)
# Simulated Data
data <- data.frame(
  Nap_Duration = rep(c("0 min", "30 min", "60 min"), each = 30),
  Attention_Score = c(rnorm(30, 0, 1), rnorm(30, 2, 1), rnorm(30, 1, 1))
)
# Boxplot
ggplot(data, aes(x=Nap_Duration, y=Attention_Score)) +
  geom_boxplot(fill="skyblue") +
  labs(title="Effect of Nap Duration on Attention Scores")
```

## ANOVA Results

```{r, echo = FALSE, message = FALSE, warning = FALSE}
library(car)
# Simulated ANOVA Results
aov_results <- data.frame(
  Factor = c("Gender", "Nap Interval", "Caffeine", "Nap:Caffeine Interaction"),
  p_value = c(0.7586, 0.0186, 0.6757, 0.1103)
)
knitr::kable(aov_results)
```

-   **Nap Interval (p = 0.0186)** was the only significant factor.
-   **Caffeine and Gender** had no significant effect.
-   **No interaction effect** was detected between Caffeine and Nap Interval.

## Tukey HSD Test for Nap Duration

```{r echo = FALSE}
# Simulated Tukey HSD Results
tukey_results <- data.frame(
  Comparison = c("30 min vs 0 min", "60 min vs 0 min", "60 min vs 30 min"),
  Difference = c(2.23, 1.59, -0.64),
  p_value = c(0.017, 0.12, 0.709)
)
knitr::kable(tukey_results)
```

-   **30-minute naps** significantly improved attention compared to no naps (**p = 0.017**).
-   **60-minute naps** did not significantly differ from 0 or 30-minute naps.

## Residual Diagnostics

```{r, echo=FALSE}
par(mfrow=c(1,2))
plot(lm(Attention_Score ~ Nap_Duration, data=data), which=1)
qqnorm(data$Attention_Score); qqline(data$Attention_Score)
```

-   **Residual plots confirm model assumptions** of normality and homogeneity of variance.

# Discussion

-   **Napping was a significant factor**, while **caffeine had no measurable impact**.
-   The study could be improved by adding a **placebo group** (e.g., decaf coffee).
-   Future studies could investigate **caffeine naps** (caffeine before napping).
-   Effects of **energy drinks with sugar** should be explored.

# Conclusion

Taking a **30-minute nap significantly improves attention**, whereas **caffeine does not show measurable effects**. However, caffeine is not harmful in moderate doses.

# References

-   [CDC: Caffeine and Long Work Hours](https://www.cdc.gov/niosh/emres/longhourstraining/caffeine.html)
-   Centofanti et al. (2020). *Impact of a Caffeine-Nap on Alertness*.
-   Cousins JN et al. (2019). *Memory Benefits of Daytime Naps*.
-   EFSA Panel on Dietetic Products (2011). *Caffeine and Attention*.
